Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-05-06 Manuscript/manuscript.Rmd
manuscript edits

**Background.** Chromosome conformation capture sequencing technologies have shown that the three-dimensional (3D) structure of the genome is folded into distinct compartments, known as topologically associated domains (TADs). The locations of TAD boundaries have been shown to be highly conserved. Additionally, the distinct patterns of epigenomic marks, binding peaks of different factors, and actively transcribed genes prompt the use of computational approaches to identify the epigenomic features that are most predictive of TAD boundaries.

@@

Commonly used measures of model performance for binary classifiers include threshold-free metrics like the area under the receiver operating curve (AUROC), and other threshold-specific metrics like accuracy, sensitivity, and specificity. The ROC curve is created by plotting the true positive rate (TPR) against the false positive rate (FPR) at various threshold settings. The subsequent AUROC, measures how well the model is capable of distinguishing between the two classes [@park2004receiver]. Accuracy is defined as the ratio of correctly identified cases (either from the positive or negative class) over the total number of cases. It measures the overall proportion of correctly classified instances. Sensitivity, also referred to as recall or TPR, is the ratio of true positives over the total number of positive classes. Specificity, on the other hand, also referred to as TNR, is the ratio of true negatives over the total number of negative classes. Each measures the proportion of correctly identified positive and negative classes respectively [@yerushalmy1947statistical]. Problems arise when using such measures in the event of imbalanced classes. For AUROC, when the prevalence of an event is low--that is, a defined genomic region truly containing a TAD boundary--mean probabilities are biased towards the majority class. This results in the area of interest under an ROC curve to be compressed toward a small corner in the ROC space [@prati2011survey]. While not directly affecting the value of the AUROC, this can have severe consequences on the interpretability of the model's performance. A better threshold-free measure to use when dealing with imbalanced data is the area under the precision-recall curve (AUPRC).  The interpretation of the AUPRC of a model is similar to that of the AUROC. However, where as the baseline model performance according to the AUROC is given by 0.50, the baseline model performance according to the AUPRC is given by the ratio of true positives over the total number of cases. This enables a more realistic benchmark to evaluate a model on as this ratio will be lower than 0.5 for a test data set with imbalanced classes. Likewise, threshold-specific metrics like accuracy and specificity are also problematic when evaluating model performance on imbalance data due to being highly influenced by the correct classification of the majority class. A better measure to consider is a composite metric involving all quadrants of the confusion matirix, such as Matthew's Correlation Coefficient (MCC) [@jeni2013facing; @chicco2017ten]. The MCC is, in essence, the correlation coefficient between the observed and predicted binary classifications [@brodersen2010balanced]. A value of positive 1 represents a perfect prediction, while a value of negative 1 represents total disagreement between prediction and observation. A value of 0 indicates no better than random prediction.

@@

In order to assess the predictive capabilities of specific types of annotations, we further compared the performances of annotation-specific models. Using the optimal set of characteristics obtained above (SMOTE re-sampling with distance-type predictors), random forest classifiers built on sets of cell line specific histone modifications, chromatin state segmentations, and transcription factor binding sites (TFBS) were compared to a model that included all three types together. For GM12878, models built simply on TFBS were found to perform consistently just as well as models that included all 3 sets of annotations, across resolution (Figure 6). Similar results were found when evaluating the K562 cell line (supplementary Figure 11). For both resolutions, models built on only histone modifications were found to have the lowest MCC, across resolutions, followed by models built on chromatin state segmentations. This suggests that TAD boundary formation is primarily carried out by transcription factor binding sites.

@@

We developed an ensemble framework to systematically compare model performances for different combinations of re-sampling and feature engineering techniques aimed at determining a consensus approach to predicting TAD boundary formation using genomic annotations. Additionally, this study provides direct comparisons of 4 different Hi-C data resolutions, and their effects on performances of downstream models. Furthermore, we go on to show the importance of using adequate metric when assessing model performance, especially in the case of highly imbalanced classes.

A typical problem facing machine learning approaches toward predicting TAD boundary formation is the introduction of highly imbalanced classes. This is because, when constructing the data for modeling, a common methodology used to determine the class of the response variable is to bin the linear genome and then create a method used to determine if each bin "includes" a TAD boundary or not. Given the sparcity of TADs throughout the genome though, this will inevitably lead to severly imbalanced classes. A workaround to this employed by [@huang2015predicting; @hong2017computational; @bednarz2014supervised] was to randomly sample from the majority class to create balanced classes in the training set. It is unclear how these studies handled

When constructing the data for modeling, a common methodology used to determine the response variable is to bin the linear genome into equally sized intervals and then create a method used to determine if each bin "includes" a TAD boundary or not. Given the sparcity of TADs throughout the genome, this will inevitably lead to severly imbalanced classes.

Although we were not able to draw direct comparisons between the GM12878 and K562 cell lines, given that they did not contain the same list of cell type specific annotations, we did find that there was relative agreement in rankings of the top predictive predictors. However, there were notable exclusions from each cell line, in addition to inconsistencies across resolutions

## LIMITATIONS

Many elastic-net models using overlap count or percent predictors were reduced to ridge, or near-ridge regression models at higher data resolutions by having the $alpha$ penalization term tuned to 0 in the cross-validation step, thereby not reducing, or reducing very few of the feature space. The results of the variable reduction step imply that far more information relating annotions to TAD boundaries is encoded in the distance-type predictors. That is, the elastic-net algorithms were better able to separate noise and reduce the feature space when performed on distance-type predictors. Likewise, this reduction was most efficient at more refined levels of data resolution.

>>>>

**Background.** Chromosome conformation capture sequencing technologies have shown that the 3-dimensional structure of the genome is folded into distinct compartments, known as topologically associated domains (TADs). The locations of TAD boundaries have been shown to be highly conserved. Additionally, the distinct patterns of epigenomic marks, binding peaks of different factors, and actively transcribed genes prompt the use of computational approaches toward identifying the epigenomic features most predictive of TAD boundaries.

@@

Commonly used measures of model performance for binary classifiers include threshold-free metrics like the area under the receiver operating curve (AUROC), and other threshold-specific metrics like accuracy, sensitivity, and specificity. The ROC curve is created by plotting the true positive rate (TPR) against the false positive rate (FPR) at various threshold settings. The subsequent AUROC, measures how well the model is capable of distinguishing between the two classes [@park2004receiver]. Accuracy is defined as the ratio of correctly identified cases (either from the positive or negative class) over the total number of cases. It measures the overall proportion of correctly classified instances. Sensitivity, also referred to as recall or TPR, is the ratio of true positives over the total number of positive classes. Specificity, on the other hand, also referred to as TNR, is the ratio of true negatives over the total number of negative classes. Each measures the proportion of correctly identified positive and negative classes respectively [@yerushalmy1947statistical]. Problems arise when using such measures in the event of imbalanced classes. For AUROC, when the prevalence of an event is low--that is, a defined genomic region truly containing a TAD boundary--mean probabilities are biased towards the majority class. This results in the area of interest under an ROC curve to be compressed toward a small corner in the ROC space [@prati2011survey; @wei2013role]. While not directly affecting the value of the AUROC, this can have severe consequences on the interpretability of the model's performance. A better threshold-free measure to use when dealing with imbalanced data is the area under the precision-recall curve (AUPRC).  The interpretation of the AUPRC of a model is similar to that of the AUROC. However, where as the baseline model performance according to the AUROC is given by 0.50, the baseline model performance according to the AUPRC is given by the ratio of true positives over the total number of cases. This enables a more realistic benchmark to evaluate a model on as this ratio will be lower than 0.5 for a test data set with imbalanced classes. Likewise, threshold-specific metrics like accuracy and specificity are also problematic when evaluating model performance on imbalance data due to being highly influenced by the correct classification of the majority class. A better measure to consider is a composite metric involving all quadrants of the confusion matirix, such as Matthew's Correlation Coefficient (MCC) [@jeni2013facing; @chicco2017ten]. The MCC is, in essence, the correlation coefficient between the observed and predicted binary classifications [@brodersen2010balanced]. A value of positive 1 represents a perfect prediction, while a value of negative 1 represents total disagreement between prediction and observation. A value of 0 indicates no better than random prediction.

@@

In order to assess the predictive capabilities of specific types of annotations, we further compared the performances of annotation-specific models. Using the optimal set of characteristics obtained above (SMOTE re-sampling with distance-type predictors), random forest classifiers built on sets of cell line specific histone modifications, chromatin state segmentations, and transcription factor binding sites (TFBS) were compared to a model that included all three types together. For GM12878, models built simply on TFBS were found to perform consistently just as well as models that included all 3 sets of annotations, across resolution (Figure 6). Similar results were found when evaluating the K562 cell line (supplementary Figure 11). For both resolutions, models built on only histone modifications were found to have the lowest MCC, across resolutions, followed by models built on chromatin state segmentations.

@@

In this study we developed an ensemble framework to systematically compare model performances for different combinations of re-sampling and feature engineering techniques aimed at determining a consensus approach to predicting TAD boundary formation using genomic annotations. We provide direct comparisons of cell line specific random forest models built of a variety of different data dependent characteristics, including 4 contact matrix resolutions ($\{\textit{10 kb}, \textit{25 kb}, \textit{50 kb}, \textit{100 kb\}}$), 3 predictor-types ($\{\textit{OC}, \textit{OP}, \textit{Distance}\}$), and 4 re-sampling techniques ($\{\textit{None}, \textit{RUS}, \textit{ROS}, \textit{SMOTE}\}$), for the GM12878 and K562 cell lines.

The experiments derived from our ensemble framework provide a useful guide toward the exploration of chromatin organization, while highlighting the impact of both class imbalance and feature engineering in TAD boundary prediction. Using the framework we demonstrated that random forest models built on imbalanced data exhibited poor performance, regardless of cell line, resolution, or predictor type. In our framework, we opted to leave the test set as a completely untouched holdout set, only performing the re-sampling, and subsequent learning, on the training set. This was done so as to incur the least amount of bias and so that all conclusions could be made soley between model characteristics. As a result, we showed that Matthew's Correlation Coefficient was much more sensitive to the imbalance present in the test set, compared to more popular metrics like AUROC, which attain artificially high values by correctly predicting majority class samples. When comparing re-sampling techniques, we found that SMOTE re-sampling outperformed models built on both simple ROS or RUS, suggesting that, in fact, a combination of both random under-sampling and random over-sampling is preferred in TAD boundary prediction. Furthermore, the most improvement was seen when combining SMOTE resampling on distance-type features when building random forest classifiers. These two characteristics exhibited consistenly better performance results across both the GM12878 and K562 cell lines. This confirmed our hypothesis that the most granular and spatially informative representation of the relationship between TAD boundaries and annotations resulted in the best performing models.

When applying SMOTE re-sampling, with distance-type predictors, on transcription factor binding sites alone, we found that random forest classifiers performed better when compared to other annotation-specific models, including histone modifications and chromatin state segmentations. Likewise the performance of TFBS specific models was comparable to models with all annotations combined, across both cell line and resolution. This suggests that the majority of the key molecular drivers aiding in the formation of TAD boundaryies are transctiption factors, and that modifications at histone tails such as acetylation or methylation play less of a role.

We also observed that, while the best set of data characteristics were consistent across resolutions, performances of classifiers built on said characteristics tended to decline as resolution decreased. For both cell lines, top performance was achieved at the 10 kb data resolution, while the worst performance was obtained at the 50 kb and 100 kb resolution for the GM12878 and K562 cell line respectively. This result can be attributed to one of two factors. First, at higher resolutions, by model construction genomic bins were smaller, Thereby resulting in more accurate profiles for the genomic annotations. Secondly, because we are mainly concerned with TAD boundaries, and not the heirarchical structure of domains as a whole, the sparcity of contact matrices at higher resolutions does not impair boundary prediction. Conversely, lower resolutions imply larger bin windows and lead to less specificity in boundary location on the linear genome, resulting in worse performances.
